---
title: "Getting started with zujson"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(zujson)
```

`zujson` exists to serve `zuhttp`: parsing a response body and building a
request body are the two things it is designed around. That is why it is
narrower than `jsonlite`, and why the type mapping below is part of the
contract rather than an implementation detail — what goes in and what comes out
should both be predictable.

Ten functions, and that is the whole package.

## Parsing

`json_parse()` accepts a character string or a raw vector; `json_parse_raw()`
and `json_parse_file()` are the explicit forms, and `json_validate()` answers
the question without building anything.

```{r}
json_parse('{"ok": true, "ids": [1, 2, 3]}')

json_validate('{"a": 1}')
json_validate("{bad")
```

JSON objects always become named lists. Arrays become an atomic vector when
their elements agree on a type and a list when they do not, so the type is
never coerced away:

```{r}
json_parse("[1, 2, 3]")
json_parse('[1, "a"]')
```

The mapping in full:

| JSON | R |
| --- | --- |
| `{"a": 1}` | `list(a = 1L)` |
| `[1, 2, 3]` | `1:3` |
| `[1, "a"]` | `list(1L, "a")` |
| `[]`, `[null, null]` | `logical(0)`, `c(NA, NA)` |
| `null` | `NULL` |

## Serializing

A **fully named** vector or list becomes an object; anything else becomes an
array. Length-1 atomic vectors unbox to bare scalars by default, so wrap one in
`I()` when a field must stay an array.

```{r}
json_write(list(a = 1, b = 2))
json_write(c(a = 1, b = 2))
json_write(list(1, 2))

json_write("x")
json_write(I("x"))
```

| R | JSON |
| --- | --- |
| `list(a = 1, b = 2)` | `{"a":1,"b":2}` |
| `c(a = 1, b = 2)` | `{"a":1,"b":2}` |
| `list(1, 2)` | `[1,2]` |
| `"x"` / `I("x")` | `"x"` / `["x"]` |
| `NA`, `NaN`, `Inf` | `null` |
| `Date`, `POSIXct` | ISO 8601 strings, UTC |
| `data.frame` | one object per row |

```{r}
json_write(list(when = as.Date("2026-09-08"), missing = NA, ratio = Inf))
```

`?json_parse` and `?json_write` carry the same tables.

## Raw bytes, both directions

A response body arrives as raw bytes and a request body should leave as raw
bytes. Going through a character string in between would mean an encoding
round trip that neither side asked for, so the raw path is direct:

```{r}
body <- json_write_raw(list(q = "cats"))
body

json_parse(body)
```

## NDJSON

NDJSON (`application/x-ndjson`) is one JSON value per line — the shape of log
tails, change feeds and bulk uploads.

```{r}
json_write_ndjson(data.frame(id = 1:2, ok = c(TRUE, FALSE)))

json_parse_ndjson('{"id":1}\n{"id":2}\n')
```

`json_write_ndjson_raw()` is the raw-bytes form.

## Safety

Parsing is meant to be pointed at an untrusted response body. yyjson validates
UTF-8, and nesting beyond 1000 levels is rejected with a structured condition
rather than walking off the C stack.

Every failure the package raises inherits from `zujson_error`, so one handler
catches all of them:

```{r}
tryCatch(json_parse("{bad"), zujson_error = function(e) conditionMessage(e))

class(tryCatch(json_parse("{bad"), zujson_error = function(e) e))
```

`zujson_info()` reports the build and the vendored yyjson version:

```{r}
zujson_info()
```

## What it deliberately does not do

No data frame reconstruction on parse, no matrix or N-d array simplification,
no incremental streaming yet, no JSON pointer or patch, no custom serializers.
`v1` is the JSON an HTTP client needs and nothing else.

## Related

`zujson` is part of the `zu*` family:
[zukomp](https://github.com/pedrobtz/zukomp) for compression,
[zuxml](https://github.com/pedrobtz/zuxml) for XML,
[zuyaml](https://github.com/pedrobtz/zuyaml) for YAML, and `zuhttp` for the
HTTP client that uses them.
