---
title: "Getting started with zujson"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(zujson)
```

`zujson` exists to serve `zuhttp`: parsing a response body and building a
request body are the two things it is designed around. That is why it is
narrower than `jsonlite`, and why the type mapping below is part of the
contract rather than an implementation detail — what goes in and what comes out
should both be predictable.

Ten functions, and that is the whole package.

## Parsing

`json_parse()` accepts a character string or a raw vector; `json_parse_raw()`
and `json_parse_file()` are the explicit forms, and `json_validate()` answers
the question without building anything.

```{r}
json_parse('{"ok": true, "ids": [1, 2, 3]}')

json_validate('{"a": 1}')
json_validate("{bad")
```

JSON objects always become named lists. Arrays become an atomic vector when
their elements agree on a type and a list when they do not, so the type is
never coerced away:

```{r}
json_parse("[1, 2, 3]")
json_parse('[1, "a"]')
```

The mapping in full:

| JSON | R |
| --- | --- |
| `{"a": 1}` | `list(a = 1L)` |
| `[1, 2, 3]` | `1:3` |
| `[1, "a"]` | `list(1L, "a")` |
| `[]`, `[null, null]` | `logical(0)`, `c(NA, NA)` |
| `null` | `NULL` |

## Simplification modes

`simplify` decides what happens to an array whose elements do not share a
kind. That is the *only* thing it changes: the promotions within the numeric
family are the same in every mode, and a nested container is never coerced
away.

```{r}
json_parse('[1, "a"]')                        # preserve, the default
json_parse('[1, "a"]', simplify = "coerce")
json_parse('[1, "a"]', simplify = "none")
```

`preserve` keeps the type and gives up the uniform shape, on the view that a
field which is usually a number and occasionally a string is a bug worth
seeing rather than one worth papering over. `coerce` is the opt-in for callers
who would rather have the vector. It follows R's own promotion exactly, so the
strings are the ones `as.character()` gives --- which matters more than it
sounds, because a hand-rolled format would not agree with R at 15 significant
digits:

```{r}
json_parse('[0.3333333333333333, "a"]', simplify = "coerce")
identical(json_parse('[1, null, "a"]', simplify = "coerce"), c("1", NA, "a"))
```

Note that JSON `null` becomes `NA` there and never the string `"NA"`.

`TRUE` and `FALSE` remain exact synonyms for `"preserve"` and `"none"`, so
nothing written against the original API changes meaning.

## Data frames

`data_frame = TRUE` turns any non-empty array whose elements are all objects
--- the shape an API uses for a list of records --- into a data frame.

```{r}
json_parse('[{"id":1,"nm":"a"},{"id":2,"nm":"b"}]', data_frame = TRUE)
```

Records are not required to agree on their keys. Columns are the union, in the
order first seen, and a record missing a key contributes `NA`, so the result is
rectangular however ragged the input:

```{r}
json_parse('[{"a":1},{"b":2},{"a":3,"c":4}]', data_frame = TRUE)
```

Each column is simplified with the active `simplify` mode, so the two features
compose rather than competing:

```{r}
str(json_parse('[{"v":1},{"v":"x"}]', data_frame = TRUE)$v)
str(json_parse('[{"v":1},{"v":"x"}]', data_frame = TRUE, simplify = "coerce")$v)
```

Anything that is not a non-empty array of objects is left exactly as it was,
and the reconstruction applies wherever such an array appears, however deeply
nested.

```{r}
str(json_parse('{"rows":[{"a":1},{"a":2}]}', data_frame = TRUE))
```

## Serializing

A **fully named** vector or list becomes an object; anything else becomes an
array. Length-1 atomic vectors unbox to bare scalars by default, so wrap one in
`I()` when a field must stay an array.

```{r}
json_write(list(a = 1, b = 2))
json_write(c(a = 1, b = 2))
json_write(list(1, 2))

json_write("x")
json_write(I("x"))
```

| R | JSON |
| --- | --- |
| `list(a = 1, b = 2)` | `{"a":1,"b":2}` |
| `c(a = 1, b = 2)` | `{"a":1,"b":2}` |
| `list(1, 2)` | `[1,2]` |
| `"x"` / `I("x")` | `"x"` / `["x"]` |
| `NA`, `NaN`, `Inf` | `null` |
| `Date`, `POSIXct` | ISO 8601 strings, UTC |
| `data.frame` | one object per row |

```{r}
json_write(list(when = as.Date("2026-09-08"), missing = NA, ratio = Inf))
```

`?json_parse` and `?json_write` carry the same tables.

## Raw bytes, both directions

A response body arrives as raw bytes and a request body should leave as raw
bytes. Going through a character string in between would mean an encoding
round trip that neither side asked for, so the raw path is direct:

```{r}
body <- json_write_raw(list(q = "cats"))
body

json_parse(body)
```

## NDJSON

NDJSON (`application/x-ndjson`) is one JSON value per line — the shape of log
tails, change feeds and bulk uploads.

```{r}
json_write_ndjson(data.frame(id = 1:2, ok = c(TRUE, FALSE)))

json_parse_ndjson('{"id":1}\n{"id":2}\n')
```

`json_write_ndjson_raw()` is the raw-bytes form.

## Safety

Parsing is meant to be pointed at an untrusted response body. yyjson validates
UTF-8, and nesting beyond 1000 levels is rejected with a structured condition
rather than walking off the C stack.

Every failure the package raises inherits from `zujson_error`, so one handler
catches all of them:

```{r}
tryCatch(json_parse("{bad"), zujson_error = function(e) conditionMessage(e))

class(tryCatch(json_parse("{bad"), zujson_error = function(e) e))
```

`zujson_info()` reports the build and the vendored yyjson version:

```{r}
zujson_info()
```

## Conformance

Claims about what a parser rejects are worth little without an external
check, so the package is run against
[JSONTestSuite](https://github.com/nst/JSONTestSuite) --- the standard
corpus for exactly this question. Its files are named by what a parser must
do: `y_` must be accepted, `n_` must be rejected, and `i_` is left to the
implementation.

```{r eval = FALSE}
# tools/jsontestsuite.R, against a pinned upstream commit
y_ must be accepted           95 files,   0 wrong
n_ must be rejected          188 files,   0 wrong
i_ implementation-defined     35 files,  12 accepted,  23 rejected
318 parsing files, 0 wrong
```

Every file is handed to `json_validate()` as **raw bytes**. Reading them as
text would be wrong twice over: the suite is full of deliberately invalid
UTF-8, which a text read mangles, and of embedded NULs, which no R string can
carry. Bytes are also what a response body actually is.

### The implementation-defined cases

The 35 `i_` files are where conforming parsers legitimately disagree, so they
are the interesting ones. The answers here follow a single split: **generous
about numbers, strict about text.**

A number outside what a `double` can hold still has a well-defined nearest
value, and R's own reader already agrees what it is --- so the document is
read and the number is converted. Refusing the whole body over one absurd
value would be a wildly disproportionate failure for an HTTP client.

```{r}
json_parse("[123123e100000]")   # overflows: Inf, as as.numeric() gives
json_parse("[123e-10000000]")   # underflows: 0
json_parse("[100000000000000000000]") # wider than int64: nearest double
```

That covers ten of the twelve acceptances: five overflow to `±Inf`, two
underflow to `0`, and three are integers too wide for `int64`.

Text is the opposite. Every string that comes back is marked as UTF-8, and
that mark is only honest because yyjson validated the bytes. Accepting
malformed text would mean handing back an R string that claims an encoding it
does not have --- so all 23 rejections are text that is not valid UTF-8,
in three groups.

Eleven involve surrogates that do not form a valid pair --- ten written as
`\uXXXX` escapes, one encoded straight into the bytes. A surrogate is only
meaningful as half of a pair, and an unpaired one encodes no character:

```{r}
json_validate('["\\ud800"]')        # lone high surrogate
json_validate('["\\uDd1e\\uD834"]') # a valid pair, inverted
json_validate('["\\ud83d\\ude02"]') # a correct pair: fine
```

Nine are invalid UTF-8 byte sequences --- truncated characters, overlong
encodings, stray continuation bytes, Latin-1 text mislabelled as JSON:

```{r}
json_validate(as.raw(c(0x22, 0xC0, 0xAF, 0x22))) # overlong encoding of "/"
json_validate(as.raw(c(0x22, 0xE9, 0x22)))       # Latin-1 e-acute, not UTF-8
```

The last three are UTF-16, with and without a byte order mark. RFC 8259 is
explicit that JSON exchanged between systems is UTF-8, so these are not JSON
this package reads; convert with `iconv()` first if you have one.

```{r}
json_validate(as.raw(c(0xFF, 0xFE, 0x22, 0x00, 0x61, 0x00, 0x22, 0x00)))
```

The remaining two acceptances are structural rather than textual. 500 nested
arrays are fine, because the depth cap is 1000 --- the cap exists to protect
the C stack, not to second-guess the sender. And a UTF-8 byte order mark is
skipped rather than rejected, because RFC 8259 forbids *emitting* one but
allows ignoring it, and real APIs emit them anyway.

```{r}
json_validate(paste0(strrep("[", 500), strrep("]", 500)))
json_parse(as.raw(c(0xEF, 0xBB, 0xBF, 0x7B, 0x7D)))  # BOM, then {}
```

## What it deliberately does not do

No matrix or N-d array simplification, no incremental streaming yet, no JSON
pointer or patch, no custom serializers. Data frame reconstruction is opt-in
rather than absent, and stays off by default so that the type mapping above is
what you get unless you ask for something else.

## How it differs from jsonlite

`jsonlite` is excellent and much broader. Everything below is the price of a
mapping small enough to hold in your head, not an oversight, and all of it is
pinned in the package's own tests --- which use `jsonlite` as an oracle over
JSON text, so a divergence that stops being deliberate fails the suite.

```{r}
library(jsonlite)
```

### Parsing

Most JSON reads the same way in both: objects, nulls, scalars, escapes,
surrogate pairs and the integer boundaries all agree. Four things do not, and
three of them have an opt-in that closes the gap.

An empty array has no elements to take a type from, so it is the zero-length
atomic vector rather than a list:

```{r}
json_parse("[]")
fromJSON("[]")
```

A mixed-kind array keeps its types unless you ask for the vector:

```{r}
json_parse('[1, "a"]')
json_parse('[1, "a"]', simplify = "coerce") # agrees with jsonlite
fromJSON('[1, "a"]')
```

An array of objects stays a list of records unless you ask for the frame:

```{r}
str(json_parse('[{"a":1},{"b":2}]'))
json_parse('[{"a":1},{"b":2}]', data_frame = TRUE) # agrees with jsonlite
```

An array of equal-length arrays is a list of rows in *every* mode. There is no
matrix detection here, so this is the one that does not converge:

```{r}
json_parse("[[1,2],[3,4]]")
fromJSON("[[1,2],[3,4]]")
```

### Serializing

This is where the two differ most, because `toJSON()` has a dozen arguments
shaping its output and `json_write()` has two.

Start with the return types, which is why the outputs below print differently:
`json_write()` gives you a plain character string, so R shows it with its
quotes escaped, while `toJSON()` gives you a classed `json` object that prints
bare. Read through the escaping when comparing the two.

The unboxing defaults are opposite --- `json_write()` unboxes a length-1
vector, `toJSON()` does not:

```{r}
json_write(35)
toJSON(35)
```

A named atomic vector is an object here and an array there. This one has no
argument that reconciles it: `toJSON()` drops vector names unless asked with
`keep_vec_names`, which it also warns about.

```{r}
json_write(c(a = 1, b = 2))
toJSON(c(a = 1, b = 2))
```

Every missing or non-finite value is `null`, where `toJSON()`'s default turns a
numeric `NA` into the *string* `"NA"` --- though a character or logical `NA`
becomes `null`, so the rule is worth knowing before you rely on it:

```{r}
json_write(c(1, NA, Inf))
toJSON(c(1, NA, Inf))
```

Doubles are written so they read back as the same number. `toJSON()` rounds to
four digits unless told otherwise:

```{r}
json_write(pi)
toJSON(pi)
```

The rest, in brief:

| R value | zujson | jsonlite, at its defaults |
| --- | --- | --- |
| `NULL` | `null` | `{}` |
| `POSIXct` | ISO 8601, always UTC | space-separated, local time |
| `matrix(1:4, 2)` | `[1,2,3,4]`, `dim` dropped | `[[1,3],[2,4]]` |
| `complex`, `raw`, `POSIXlt` | `zujson_unsupported_type` | `"2+2i"`, `"Ymxh"`, a string |
| a data frame column that is itself a data frame | `zujson_unsupported_type` | nested objects |

### Failing

The divergence that matters most to a client. Every failure here is a typed
condition rather than a bare `simpleError`, so one handler catches the lot and
can tell the cases apart:

```{r}
class(tryCatch(json_parse("{bad"), error = function(e) e))
class(tryCatch(fromJSON("{bad"), error = function(e) e))
```

Two inputs are refused on purpose. An escaped NUL is valid JSON that no R
string can hold, so `json_parse()` raises; `fromJSON()` silently returns a
truncated string instead:

```{r}
nul <- '["a\\u0000b"]'
tryCatch(json_parse(nul), zujson_error = function(e) conditionMessage(e))
fromJSON(nul)
```

And nesting past 1000 levels is a `zujson_depth_error` rather than a C stack
overflow --- which is what makes the parser safe to point at a body you did not
write.

Finally, `json_parse()` never performs I/O. `fromJSON()` inspects its argument
and will fetch a URL or read a file path if it looks like one; here reading a
file is `json_parse_file()`, and it is always explicit.

## Related

`zujson` is part of the `zu*` family:
[zukomp](https://github.com/pedrobtz/zukomp) for compression,
[zuxml](https://github.com/pedrobtz/zuxml) for XML,
[zuyaml](https://github.com/pedrobtz/zuyaml) for YAML, and `zuhttp` for the
HTTP client that uses them.
