---
title: "Getting started with zujson"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(zujson)
```

`zujson` exists to serve `zuhttp`: parsing a response body and building a
request body are the two things it is designed around. That is why it is
narrower than `jsonlite`, and why the type mapping below is part of the
contract rather than an implementation detail — what goes in and what comes out
should both be predictable.

Ten functions, and that is the whole package.

## Parsing

`json_parse()` accepts a character string or a raw vector; `json_parse_raw()`
and `json_parse_file()` are the explicit forms, and `json_validate()` answers
the question without building anything.

```{r}
json_parse('{"ok": true, "ids": [1, 2, 3]}')

json_validate('{"a": 1}')
json_validate("{bad")
```

JSON objects always become named lists. Arrays become an atomic vector when
their elements agree on a type and a list when they do not, so the type is
never coerced away:

```{r}
json_parse("[1, 2, 3]")
json_parse('[1, "a"]')
```

The mapping in full:

| JSON | R |
| --- | --- |
| `{"a": 1}` | `list(a = 1L)` |
| `[1, 2, 3]` | `1:3` |
| `[1, "a"]` | `list(1L, "a")` |
| `[]`, `[null, null]` | `logical(0)`, `c(NA, NA)` |
| `null` | `NULL` |

## Simplification modes

`simplify` decides what happens to an array whose elements do not share a
kind. That is the *only* thing it changes: the promotions within the numeric
family are the same in every mode, and a nested container is never coerced
away.

```{r}
json_parse('[1, "a"]')                        # preserve, the default
json_parse('[1, "a"]', simplify = "coerce")
json_parse('[1, "a"]', simplify = "none")
```

`preserve` keeps the type and gives up the uniform shape, on the view that a
field which is usually a number and occasionally a string is a bug worth
seeing rather than one worth papering over. `coerce` is the opt-in for callers
who would rather have the vector. It follows R's own promotion exactly, so the
strings are the ones `as.character()` gives --- which matters more than it
sounds, because a hand-rolled format would not agree with R at 15 significant
digits:

```{r}
json_parse('[0.3333333333333333, "a"]', simplify = "coerce")
identical(json_parse('[1, null, "a"]', simplify = "coerce"), c("1", NA, "a"))
```

Note that JSON `null` becomes `NA` there and never the string `"NA"`.

`TRUE` and `FALSE` remain exact synonyms for `"preserve"` and `"none"`, so
nothing written against the original API changes meaning.

## Data frames

`data_frame = TRUE` turns any non-empty array whose elements are all objects
--- the shape an API uses for a list of records --- into a data frame.

```{r}
json_parse('[{"id":1,"nm":"a"},{"id":2,"nm":"b"}]', data_frame = TRUE)
```

Records are not required to agree on their keys. Columns are the union, in the
order first seen, and a record missing a key contributes `NA`, so the result is
rectangular however ragged the input:

```{r}
json_parse('[{"a":1},{"b":2},{"a":3,"c":4}]', data_frame = TRUE)
```

Each column is simplified with the active `simplify` mode, so the two features
compose rather than competing:

```{r}
str(json_parse('[{"v":1},{"v":"x"}]', data_frame = TRUE)$v)
str(json_parse('[{"v":1},{"v":"x"}]', data_frame = TRUE, simplify = "coerce")$v)
```

Anything that is not a non-empty array of objects is left exactly as it was,
and the reconstruction applies wherever such an array appears, however deeply
nested.

```{r}
str(json_parse('{"rows":[{"a":1},{"a":2}]}', data_frame = TRUE))
```

## Serializing

A **fully named** vector or list becomes an object; anything else becomes an
array. Length-1 atomic vectors unbox to bare scalars by default, so wrap one in
`I()` when a field must stay an array.

```{r}
json_write(list(a = 1, b = 2))
json_write(c(a = 1, b = 2))
json_write(list(1, 2))

json_write("x")
json_write(I("x"))
```

| R | JSON |
| --- | --- |
| `list(a = 1, b = 2)` | `{"a":1,"b":2}` |
| `c(a = 1, b = 2)` | `{"a":1,"b":2}` |
| `list(1, 2)` | `[1,2]` |
| `"x"` / `I("x")` | `"x"` / `["x"]` |
| `NA`, `NaN`, `Inf` | `null` |
| `Date`, `POSIXct` | ISO 8601 strings, UTC |
| `data.frame` | one object per row |

```{r}
json_write(list(when = as.Date("2026-09-08"), missing = NA, ratio = Inf))
```

`?json_parse` and `?json_write` carry the same tables.

## Raw bytes, both directions

A response body arrives as raw bytes and a request body should leave as raw
bytes. Going through a character string in between would mean an encoding
round trip that neither side asked for, so the raw path is direct:

```{r}
body <- json_write_raw(list(q = "cats"))
body

json_parse(body)
```

## NDJSON

NDJSON (`application/x-ndjson`) is one JSON value per line — the shape of log
tails, change feeds and bulk uploads.

```{r}
json_write_ndjson(data.frame(id = 1:2, ok = c(TRUE, FALSE)))

json_parse_ndjson('{"id":1}\n{"id":2}\n')
```

`json_write_ndjson_raw()` is the raw-bytes form.

## Safety

Parsing is meant to be pointed at an untrusted response body. yyjson validates
UTF-8, and nesting beyond 1000 levels is rejected with a structured condition
rather than walking off the C stack.

Every failure the package raises inherits from `zujson_error`, so one handler
catches all of them:

```{r}
tryCatch(json_parse("{bad"), zujson_error = function(e) conditionMessage(e))

class(tryCatch(json_parse("{bad"), zujson_error = function(e) e))
```

`zujson_info()` reports the build and the vendored yyjson version:

```{r}
zujson_info()
```

## Conformance

Claims about what a parser rejects are worth little without an external
check, so the package is run against
[JSONTestSuite](https://github.com/nst/JSONTestSuite) --- the standard
corpus for exactly this question. Its files are named by what a parser must
do: `y_` must be accepted, `n_` must be rejected, and `i_` is left to the
implementation.

```{r eval = FALSE}
# tools/jsontestsuite.R, against a pinned upstream commit
y_ must be accepted           95 files,   0 wrong
n_ must be rejected          188 files,   0 wrong
i_ implementation-defined     35 files,  12 accepted,  23 rejected
318 parsing files, 0 wrong
```

Every file is handed to `json_validate()` as **raw bytes**. Reading them as
text would be wrong twice over: the suite is full of deliberately invalid
UTF-8, which a text read mangles, and of embedded NULs, which no R string can
carry. Bytes are also what a response body actually is.

### The implementation-defined cases

The 35 `i_` files are where conforming parsers legitimately disagree, so they
are the interesting ones. The answers here follow a single split: **generous
about numbers, strict about text.**

A number outside what a `double` can hold still has a well-defined nearest
value, and R's own reader already agrees what it is --- so the document is
read and the number is converted. Refusing the whole body over one absurd
value would be a wildly disproportionate failure for an HTTP client.

```{r}
json_parse("[123123e100000]")   # overflows: Inf, as as.numeric() gives
json_parse("[123e-10000000]")   # underflows: 0
json_parse("[100000000000000000000]") # wider than int64: nearest double
```

That covers ten of the twelve acceptances: five overflow to `±Inf`, two
underflow to `0`, and three are integers too wide for `int64`.

Text is the opposite. Every string that comes back is marked as UTF-8, and
that mark is only honest because yyjson validated the bytes. Accepting
malformed text would mean handing back an R string that claims an encoding it
does not have --- so all 23 rejections are text that is not valid UTF-8,
in three groups.

Eleven involve surrogates that do not form a valid pair --- ten written as
`\uXXXX` escapes, one encoded straight into the bytes. A surrogate is only
meaningful as half of a pair, and an unpaired one encodes no character:

```{r}
json_validate('["\\ud800"]')        # lone high surrogate
json_validate('["\\uDd1e\\uD834"]') # a valid pair, inverted
json_validate('["\\ud83d\\ude02"]') # a correct pair: fine
```

Nine are invalid UTF-8 byte sequences --- truncated characters, overlong
encodings, stray continuation bytes, Latin-1 text mislabelled as JSON:

```{r}
json_validate(as.raw(c(0x22, 0xC0, 0xAF, 0x22))) # overlong encoding of "/"
json_validate(as.raw(c(0x22, 0xE9, 0x22)))       # Latin-1 e-acute, not UTF-8
```

The last three are UTF-16, with and without a byte order mark. RFC 8259 is
explicit that JSON exchanged between systems is UTF-8, so these are not JSON
this package reads; convert with `iconv()` first if you have one.

```{r}
json_validate(as.raw(c(0xFF, 0xFE, 0x22, 0x00, 0x61, 0x00, 0x22, 0x00)))
```

The remaining two acceptances are structural rather than textual. 500 nested
arrays are fine, because the depth cap is 1000 --- the cap exists to protect
the C stack, not to second-guess the sender. And a UTF-8 byte order mark is
skipped rather than rejected, because RFC 8259 forbids *emitting* one but
allows ignoring it, and real APIs emit them anyway.

```{r}
json_validate(paste0(strrep("[", 500), strrep("]", 500)))
json_parse(as.raw(c(0xEF, 0xBB, 0xBF, 0x7B, 0x7D)))  # BOM, then {}
```

## What it deliberately does not do

No matrix or N-d array simplification, no incremental streaming yet, no JSON
pointer or patch, no custom serializers. Data frame reconstruction is opt-in
rather than absent, and stays off by default so that the type mapping above is
what you get unless you ask for something else.

## How it differs from jsonlite

`jsonlite` is excellent and much broader. Everything below is the price of a
mapping small enough to hold in your head, not an oversight, and all of it is
pinned in the package's own tests --- which use `jsonlite` as an oracle over
JSON text, so a divergence that stops being deliberate fails the suite.

Every cell in the tables below is computed by running both packages when this
page is built, so none of it can quietly go stale.

```{r include = FALSE}
library(jsonlite)

# One compact line per R value. Data frames and matrices are rebuilt as the
# call that would make them, which is shorter than deparsing the attributes
# and still says exactly what came back.
fmt <- function(x) {
  if (is.data.frame(x)) {
    cols <- vapply(x, function(col) paste(deparse(col), collapse = ""), "")
    return(sprintf("data.frame(%s)",
                   paste(names(x), cols, sep = " = ", collapse = ", ")))
  }
  if (is.matrix(x)) {
    return(sprintf("matrix(%s, nrow = %d)",
                   paste(deparse(as.vector(x)), collapse = ""), nrow(x)))
  }
  paste(deparse(x, width.cutoff = 500), collapse = " ")
}
code <- function(x) paste0("`", x, "`")

# Evaluating the same expression that gets printed as the label is what keeps
# the two columns honest: there is no second copy to fall out of step.
run <- function(e, f) {
  tryCatch(f(eval(e)), zujson_error = function(c) paste0("error: ", class(c)[1]),
           error = function(c) paste0("error: ", class(c)[1]))
}
label <- function(e) paste(deparse(e), collapse = "")
```

### Parsing

Most JSON reads the same way in both: objects, nulls, scalars, escapes,
surrogate pairs and the integer boundaries all agree. Four things do not, and
three of them have an opt-in that closes the gap.

```{r echo = FALSE}
js <- c("[]", '[1, "a"]', '[{"a":1},{"b":2}]', "[[1,2],[3,4]]")
knitr::kable(data.frame(
  JSON = code(js),
  zujson = code(vapply(js, function(j) fmt(json_parse(j)), "")),
  jsonlite = code(vapply(js, function(j) fmt(fromJSON(j)), "")),
  `Opt-in that agrees` = c("none needed --- see below",
                           '`simplify = "coerce"`',
                           "`data_frame = TRUE`",
                           "none; no matrix detection"),
  check.names = FALSE, row.names = NULL
), row.names = FALSE)
```

The empty array is the one with no opt-in and no argument about it: with no
elements to take a type from, the zero-length atomic vector is as good an
answer as the empty list, and this package prefers the one that keeps `[]` and
`[1,2]` in the same family.

### Serializing

This is where the two differ most, because `toJSON()` has a dozen arguments
shaping its output and `json_write()` has two.

```{r echo = FALSE}
es <- list(
  quote(35),
  quote(c(a = 1, b = 2)),
  quote(c(1, NA, Inf)),
  quote(pi),
  quote(NULL),
  quote(as.POSIXct("2013-06-17 22:33:44", tz = "UTC")),
  quote(matrix(1:4, 2)),
  quote(complex(real = 2, imaginary = 2)),
  quote(charToRaw("bla")),
  quote(as.POSIXlt("2013-06-17 22:33:44", tz = "UTC"))
)
knitr::kable(data.frame(
  `R value` = code(vapply(es, label, "")),
  zujson = code(vapply(es, run, "", f = json_write)),
  `jsonlite, at its defaults` =
    code(vapply(es, run, "", f = function(x) as.character(toJSON(x)))),
  check.names = FALSE, row.names = NULL
), row.names = FALSE)
```

Four of those are worth a sentence.

The unboxing defaults are **opposite**: `json_write()` unboxes a length-1
vector and `toJSON()` does not, which is the first thing to trip over when
porting code. A **named atomic vector** is an object here and an array there,
and this is the one with no argument that reconciles it --- `toJSON()` drops
vector names unless asked with `keep_vec_names`, which it also warns about.
Every **missing or non-finite** value is `null` here, where `toJSON()`'s
default makes a numeric `NA` the *string* `"NA"` while a character or logical
`NA` becomes `null`. And **doubles** are written so they read back as the same
number, rather than rounded to four digits.

The refusals at the bottom of the table are deliberate: a complex number, a
raw vector and a `POSIXlt` have no obvious JSON form, so they raise
`zujson_unsupported_type` instead of being guessed at. A data frame column that
is itself a data frame is refused for the same reason. `POSIXct` is always UTC,
which is the same instant whatever `tzone` says.

### Failing

The divergence that matters most to a client: every failure here is a typed
condition rather than a bare `simpleError`, so one handler catches the lot and
can still tell the cases apart.

```{r echo = FALSE}
cls <- function(e) paste(class(tryCatch(eval(e), error = function(c) c))[1])
deep <- paste0(strrep("[", 1500), strrep("]", 1500))
nul <- '["a\\u0000b"]'
knitr::kable(data.frame(
  Input = c("`\"{bad\"`", "an escaped NUL, `\"\\u0000\"`",
            "nesting 1500 deep", "a URL or file path"),
  zujson = c(code(cls(quote(json_parse("{bad")))),
             code(cls(quote(json_parse(nul)))),
             code(cls(quote(json_parse(deep)))),
             "never read; `json_parse_file()` is explicit"),
  jsonlite = c(code(cls(quote(fromJSON("{bad")))),
               paste0("silently truncates to ", code(fromJSON(nul))),
               "C stack overflow",
               "fetched or read"),
  check.names = FALSE, row.names = NULL
), row.names = FALSE)
```

The middle two are the ones to notice. An escaped NUL is valid JSON that no R
string can hold, so refusing it is the only honest answer --- returning a
silently shortened string loses data without saying so. And the depth cap turns
what would be a C stack overflow into a condition you can catch, which is what
makes the parser safe to point at a body you did not write.

## Related

`zujson` is part of the `zu*` family:
[zukomp](https://github.com/pedrobtz/zukomp) for compression,
[zuxml](https://github.com/pedrobtz/zuxml) for XML,
[zuyaml](https://github.com/pedrobtz/zuyaml) for YAML, and `zuhttp` for the
HTTP client that uses them.
